Detection of Oscillations in Aldebaran with Ground-Based Observations

### 2018-02-02 paper/aldebaran.tex
Minor changes with a noteable note on RGB only treatment.

\begin{abstract}
The nearby red giant Aldebaran is known to host a gas giant planetary companion from decades of ground-based spectroscopic measurements of its Doppler shift. Using Gaussian Process-based Continuous Auto-Regressive Moving Average (CARMA) models, we show that these historic data also contain evidence of acoustic oscillations in the star itself, and verify this result with further dedicated ground-based spectroscopy and space-based photometry with the \kepler Space Telescope. From the frequency of these oscillations we determine the mass of Aldebaran to be $1.16 \pm 0.07$~\msun, and note that this implies its planet will have been subject to insolation comparable to the Earth for some of the star's main sequence lifetime. Our approach to sparse, irregularly sampled time series astronomical observations has the potential to unlock similar measurements for hundreds of stars in archival data. %, and permits more flexible observing schedules in future.
\end{abstract}

@@

In this paper, we present a re-analysis of these original RV data in which we not only confirm the planetary signal, but detect acoustic oscillations in Aldebaran for the first time. \citet{hatzes1993} noted night-to-night RV variability: in this planet-hunting noise we find an asteroseismic signal. We validate this method and its result with new independent RV observations with the Hertzsprung SONG~Telescope, and photometry from the \ktwo Mission. By measuring the frequency of maximum power of these \emph{p}-mode oscillations, \numax, we asteroseismically determine the mass of Aldebaran to be $1.16 \pm 0.07$~\msun. This precise stellar mass allows us to calculate that Aldebaran~b and any satellites it may have, although they are now likely to be very hot, would have had equilibrium temperatures comparable to that of the Earth when Aldebaran was on the main sequence. It is possible that they may have once been habitable, billions of years ago.

@@

%Stellar properties can be determined with the average seismic measures and effective temperature \teff.  A set of scaling relations exist that illustrate the constraint the global asteroseismic measures provide.  These scaling relations are understood to be approximate but never the less are commonly used for radius $R$, mass $M$, and surface gravity $g$:
These parameters can be used to estimate the radius $R$, mass $M$, and surface gravity $g$ of stars when combined with an estimate of the effective temperature \teff\ through so-called scaling relations:

@@

While the accuracy of these scaling relations is still a matter of ongoing work \citep[e.g.,][]{2017ApJ...844..102H}, stellar parameters can be estimated by comparing observables to parameter from models of stellar evolution.

@@

\subsection{SONG Observations}    % this goes into section 4.1
\label{sec:SONG-observations} In order to confirm the oscillations detected in
the archival data, we used the Hertzsprung SONG telescope
\citep{2017ApJ...836..142G} to conduct high time cadence follow-up observations.
These were carried out in the highest resolution mode ($R\approx110000$) with an
integration time of 30\,s, between 2016 September 27 and 2017 January 12. During
this time we attempted to obtain at least one visit per available night, for a
total of 254 epochs over the campaign. The spectra were extracted using the SONG
pipeline \citep[see][]{2017ApJ...836..142G}. SONG employs an iodine cell for
precise wavelength calibration and the velocities were determined using the
\textit{iSONG} software following the same procedures as described in
\cite{2017ApJ...836..142G}. The typical velocity precision per visit is
2.5$\,\rm m\,s^{-1}$ allowing us to easily detect the oscillations which
displays a peak-to-peak amplitude of $\approx \, 170 \, \rm m \, s^{-1}$.
Figure~\ref{songdata} displays the radial velocities obtained -- oscillations
are easily visible as well as a long-term trend, probably due to the planetary
companion.  From this data set alone we also find evidence for a low-quality ($Q
= \SONGQRange{}$) oscillatory mode with $\nu_\mathrm{max} = \SONGNuMaxRange{} \,
\mu\mathrm{Hz}$ which is consistent with the inference from the archival data.
This posterior is shown in Figure \ref{fig:numax-datasets}.

@@

Figure \ref{k2_lightcurve} shows the light curve and power spectrum of the K2 observations of Aldebaran. We detect clear variability on a average timescale of $\sim$\,5\,days, consistent with the expected timescale for solar-like oscillations for a high-luminosity red giant. To measure global asteroseismic parameters we model the background variability in the Fourier domain using the methodology described in \citet{huber09}, yielding a frequency of maximum power of $\numax = 2.2 \pm 0.25\,\muHz$ and amplitude per radial mode of $1850 \pm 500$\,ppm. Due to the limited frequency resolution of the 70-day K2 timeseries we were unable to measure the large frequency separation \Dnu, which is expected to be on the order of $\approx$\,0.5\muHz.

@@

The stellar oscillations are best fit with a single, low-quality ($Q =
\CombinedQRange{}$) oscillatory mode with $\numax = 2.238^{+ 0.086}_{- 0.080}$
\muHz.  We find no improvement in the marginal likelihood (evidence) or other
model-selection information criteria \citep{Gelman2013} from models with
additional oscillatory modes, so we conclude that the large spacing $\Delta\nu$
is not observable in this data set.  We have determined the stellar properties
of Aldebaran with the additional constraint of \numax determined from this
analysis.  The details of the stellar modelling are presented in Appendix
\ref{sec:mod}.  We have considered the impact of our additional \numax
constraint and run our analysis for different sets of spectroscopic estimates.
For final values we adopt the \citet{2012Sheffield} spectroscopic solution as
being `middle of the road' estimates.  For the analysis without constraint on
\numax we find estimates of the stellar properties as $M = 1.27^{+0.24}_{-0.2}
\, \mathrm{M_{\odot}}$ and age $4.86^{+3.56}_{-2.04} \, \rm Gyr$.  With the
inclusion of \numax we find $M = 1.16^{+0.07}_{-0.07} \, \mathrm{M_{\odot}}$ and
age $6.38^{+1.42}_{-1.12} \, \rm Gyr$.  It is clear that the addition of \numax
provides substantially more precise estimates of mass and age. With this stellar
mass, the radial velocity signal translates to an inclination-dependent
planetary mass of $M\,\sin\,{i} = 5.81 \pm 0.72~M_J$
\citep[Equation~1]{Torres2008}, a massive giant planet; for sufficiently high
inclinations ($\gtrsim 45^\circ$), the mass of Aldebaran~b may exceed $13~M_J$,
making it a possible low-mass brown dwarf candidate. A recent study by \citet{2018arXiv180105239H} has cast doubt on the validity of the detection of the massive planet $\gamma$~Draconis~b, and by extension similar planets around similar K~giants such as Aldebaran. We remain convinced that the $\gamma$~Draconis problem is not an issue in this case, and Aldebaran~b is a bona fide planet, for reasons outlined in Appendix~\ref{gamma_dra}.

@@

\section{Stellar Modelling}\label{sec:mod}
\subsection{Stellar Models}\label{sec:stell_mod}
We used our determination of \numax and several combinations of the asteroseismic and spectroscopic parameters, along with luminosity, to estimate the fundamental stellar parameters, via fitting to stellar models. We used \textsc{MESA} models \citep{2011Paxton,2013Paxton} in conjunction with the Bayesian code \textsc{PARAM} \citep{2006dasilva, 2017Rod}.  A summary of our selected ``benchmark'' options is as follows;
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}%GUY THIS BIT COULD PROBABLY BE DROPPED AND REFERENCED OUT TO \citep{2017Rod}
\item Heavy element partitioning from \cite{1993Grevesse}.
\item OPAL equation of state \citep{2002Rogers} along with OPAL opacities \citep{1996Iglesias}, with complementary values at low temperatures from \cite{2005Ferguson}.
\item Nuclear reaction rates from NACRE \citep{1999Angulo}.
\item The atmosphere model according to \cite{1966Kris}.
\item The mixing length theory to describe convection (we adopt a solar-calibrated parameter $\alpha_{\textrm{MLT}} =1.9657$).
\item Convective overshooting on the main sequence is set to $\alpha_{\textrm{ov}}=0.2H_{p}$, with $H_{p}$ the pressure scale height at the border of the convective core. Overshooting was applied according to the \cite{1975Maeder} step function
scheme.
\item No rotational mixing or diffusion is included.
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We do not need to correct for the line-of-sight Doppler shift at the frequency precision available in our data \citep{2014MNRAS.445L..94D}.
\subsection{Additional modelling inputs}\label{sec:addmod}
In addition to the asteroseismic parameters, spectroscopically-determined temperature and metallicity values are needed. There exist multiple literature values for Aldebaran. We chose to compare a range of literature values to investigate what uncertainty these systematically-differing models produce in inferred stellar properties.

>>>>

\begin{abstract}
The nearby red giant Aldebaran is known to host a gas giant planetary companion from decades of ground-based spectroscopic measurements of its Doppler shift. Using Gaussian Process-based Continuous Auto-Regressive Moving Average (CARMA) models, we show that these historic data also contain evidence of acoustic oscillations in the star itself, and verify this result with further dedicated ground-based spectroscopy and space-based photometry with the \kepler Space Telescope. From the frequency of these oscillations we determine the mass of Aldebaran to be $1.16 \pm 0.07$~\msun, and note that this implies its planet will have been subject to insolation comparable to the Earth for some of the star's main sequence lifetime. Our approach to sparse, irregularly sampled time series astronomical observations has the potential to unlock asteroseismic measurements for thousands of stars in archival data, and push to lower-mass planets around red giant stars.
\end{abstract}

@@

In this paper, we present a re-analysis of these original RV data in which we not only confirm the planetary signal, but detect acoustic oscillations in Aldebaran for the first time. \citet{hatzes1993} noted night-to-night RV variability: in this planet-hunting noise we find an asteroseismic signal. We validate this method and its result with new independent RV observations with the Hertzsprung SONG~Telescope, and photometry from the \ktwo Mission. By measuring the frequency of maximum power of these \emph{p}-mode oscillations, \numax, we can determine the mass of Aldebaran to be $1.16 \pm 0.07$~\msun. This precise stellar mass allows us to calculate that Aldebaran~b and any satellites it may have, although they are now likely to be very hot, would have had equilibrium temperatures comparable to that of the Earth when Aldebaran was on the main sequence. It is possible that they may have once been habitable, billions of years ago.

@@

%Stellar properties can be determined with the average seismic measures and effective temperature \teff.  A set of scaling relations exist that illustrate the constraint the global asteroseismic measures provide.  These scaling relations are understood to be approximate but never the less are commonly used for radius $R$, mass $M$, and surface gravity $g$:
These parameters can be used to estimate the radius $R$, mass $M$, and surface gravity $g$ of stars when combined with an estimate of the effective temperature \teff\ through scaling relations:

@@

While the accuracy of these scaling relations is still a matter of ongoing work \citep[e.g.,][]{2017ApJ...844..102H}, stellar parameters can be estimated by comparing observables to parameters from models of stellar evolution.

@@

\subsection{SONG Observations}    % this goes into section 4.1
\label{sec:SONG-observations} In order to confirm the oscillations detected in
the archival data, we used the Hertzsprung SONG telescope
\citep{2017ApJ...836..142G} to conduct high time cadence follow-up observations.
These were carried out in the highest resolution mode ($R\approx110000$) with an
integration time of 30\,s, between 2016 September 27 and 2017 January 12. During
this time we attempted to obtain at least one visit per available night, for a
total of 254 epochs over the campaign. The spectra were extracted using the SONG
pipeline \citep[see][]{2017ApJ...836..142G}. SONG employs an iodine cell for
precise wavelength calibration and the velocities were determined using the
\textit{iSONG} software following the same procedures as described in
\cite{2017ApJ...836..142G}. The typical velocity precision per visit is
2.5$\,\rm m\,s^{-1}$ allowing us to easily detect the oscillations which
display a peak-to-peak amplitude of $\approx \, 170 \, \rm m \, s^{-1}$.
Figure~\ref{songdata} displays the radial velocities obtained -- oscillations
are easily visible as well as a long-term trend, probably due to the planetary
companion.  From this data set alone we also find evidence for a low-quality ($Q
= \SONGQRange{}$) oscillatory mode with $\nu_\mathrm{max} = \SONGNuMaxRange{} \,
\mu\mathrm{Hz}$ which is consistent with the inference from the archival data.
This posterior is shown in Figure \ref{fig:numax-datasets}.

@@

Figure \ref{k2_lightcurve} shows the light curve and power spectrum of the K2 observations of Aldebaran. We detect clear variability on an average timescale of $\sim$\,5\,days, consistent with the expected timescale for solar-like oscillations for a high-luminosity red giant. To measure global asteroseismic parameters we model the background variability in the Fourier domain using the methodology described in \citet{huber09}, yielding a frequency of maximum power of $\numax = 2.2 \pm 0.25\,\muHz$ and amplitude per radial mode of $1850 \pm 500$\,ppm. Due to the limited frequency resolution of the 70-day K2 timeseries we were unable to measure the large frequency separation \Dnu, which is expected to be on the order of $\approx$\,0.5\muHz.

@@

The stellar oscillations are best fit with a single, low-quality ($Q =
\CombinedQRange{}$) oscillatory mode with $\numax = 2.238^{+ 0.086}_{- 0.080}$
\muHz.  We find no improvement in the marginal likelihood (evidence) or other
model-selection information criteria \citep{Gelman2013} from models with
additional oscillatory modes, so we conclude that the large spacing $\Delta\nu$
is not observable in this data set.

We have determined the stellar properties
of Aldebaran with the additional constraint of \numax determined from this
analysis.  The details of the stellar modelling are presented in Appendix
\ref{sec:mod}.  We have considered the impact of our additional \numax
constraint and run our analysis for different sets of spectroscopic estimates.
For final values we adopt the \citet{2012Sheffield} spectroscopic solution as
being `middle of the road' estimates.  For the analysis without constraint on
\numax we find estimates of the stellar properties as $M = 1.27^{+0.24}_{-0.2}
\, \mathrm{M_{\odot}}$ and age $4.86^{+3.56}_{-2.04} \, \rm Gyr$.  With the
inclusion of \numax we find $M = 1.16^{+0.07}_{-0.07} \, \mathrm{M_{\odot}}$ and
age $6.38^{+1.42}_{-1.12} \, \rm Gyr$.  It is clear that the addition of \numax
provides substantially more precise estimates of mass and age. With this stellar
mass, the radial velocity signal translates to an inclination-dependent
planetary mass of $M\,\sin\,{i} = 5.81 \pm 0.72~M_J$
\citep[Equation~1]{Torres2008}, a massive giant planet; for sufficiently high
inclinations ($\gtrsim 45^\circ$), the mass of Aldebaran~b may exceed $13~M_J$,
making it a possible low-mass brown dwarf candidate. A recent study by \citet{2018arXiv180105239H} has cast doubt on the validity of the detection of the massive planet $\gamma$~Draconis~b, and by extension similar planets around similar K~giants such as Aldebaran. We remain convinced that the $\gamma$~Draconis problem is not an issue in this case, and Aldebaran~b is a bona fide planet, for reasons outlined in Appendix~\ref{gamma_dra}.

@@

\section{Stellar Modelling}\label{sec:mod}
\subsection{Stellar Models}\label{sec:stell_mod}
We used our determination of \numax and several combinations of the asteroseismic and spectroscopic parameters, along with luminosity, to estimate the fundamental stellar parameters, via fitting to stellar models. We used \textsc{MESA} models \citep{2011Paxton,2013Paxton} in conjunction with the Bayesian code \textsc{PARAM} \citep{2006dasilva, 2017Rod}.  A summary of our selected ``benchmark'' options is as follows;
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}%GUY THIS BIT COULD PROBABLY BE DROPPED AND REFERENCED OUT TO \citep{2017Rod}
\item Heavy element partitioning from \cite{1993Grevesse}.
\item OPAL equation of state \citep{2002Rogers} along with OPAL opacities \citep{1996Iglesias}, with complementary values at low temperatures from \cite{2005Ferguson}.
\item Nuclear reaction rates from NACRE \citep{1999Angulo}.
\item The atmosphere model according to \cite{1966Kris}.
\item The mixing length theory to describe convection (we adopt a solar-calibrated parameter $\alpha_{\textrm{MLT}} =1.9657$).
\item Convective overshooting on the main sequence is set to $\alpha_{\textrm{ov}}=0.2H_{p}$, with $H_{p}$ the pressure scale height at the border of the convective core. Overshooting was applied according to the \cite{1975Maeder} step function
scheme.
\item We consider only the red giant branch phase to avoid the treatment of poorly understood mass loss at the tip of the red giant branch.
\item No rotational mixing or diffusion is included.
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We do not need to correct for the line-of-sight Doppler shift at the frequency precision available in our data \citep{2014MNRAS.445L..94D}.
\subsection{Additional modelling inputs}\label{sec:addmod}
In addition to the asteroseismic parameters, spectroscopically-determined temperature and metallicity values are needed. There exist multiple literature values for Aldebaran. We chose to compare a range of literature values to investigate what uncertainty these systematically-differing models produce in inferred stellar properties.
